Data Migration under a Changing Schema in Ampersand

### 2023-09-30 2022Migration/articleMigrationFACS.tex
minor
WIP on the introduction

\usepackage{hyperref}
%\usepackage{multicol}
%\usepackage{footmisc}
%\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amssymb,stmaryrd}
\usepackage{mathrsfs}
\usepackage{mathtools}
%\usepackage{amsthm}
\usepackage[english]{babel}
%\usepackage[official,right]{eurosym}
\selectlanguage{english}
\hyphenation{ExecEngine}
%\newtheorem{lemma}{Lemma}

@@

\title{Data Migration under a Changing Schema}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Sebastiaan Joosten\inst{1}\orcidID{0000-0002-6590-6220}\\ \and
Stef Joosten\inst{2,3}\orcidID{0000-0001-8308-0189}}
%
\authorrunning{S. Joosten}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{University of Minnesota, Minneapolis, USA \and Open Universiteit Nederland, Heerlen, the Netherlands\\ 
\email{stef.joosten@ou.nl} \and
Ordina NV, Nieuwegein, the Netherlands}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Software generators can help increase the frequency of releases and their reliability.
They save on time spent on development and fixing human-induced mistakes by compiling a specification into a working information system.
However, many generators do not support data migrations.
Data migration is necessary when an incremental deployment changes the system's schema.
Consequently, developers tend to avoid migrations or migrate data ``by hand''.

@@

\keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change.}
\end{abstract}
%
%
%
\section{Introduction}
\label{sct:Introduction}
   The purpose of this work is to automate the data migrations that come with incrementally developing and deploying information systems.
   This paper proposes a theory for such data migrations in the context of generated information systems.
   In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.

   A problem is that software generators typically do not support the data migration process, especially when the schema changes.
   The ``manual'' effort required to migrate data slows down the development process and introduces mistakes.
   Such effects can be measured in terms of DevOps metrics such as deployment frequency, reliability of deployments, and change failure rate~\cite{DevOps2021}.

   This paper proposes a theory for data migrations that is meant to be implemented in a software generator called Ampersand~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018}.
   It aims at automating data migrations, so incremental changes can be done more reliably and faster.
   This contributes to a more agile software development process.
   We believe that users will experience a more organic evolution of the system because increments are smaller, occur more frequently and more reliably.

   This paper assumes that the existing system and the desired system have different schemas.
   The existing system is the system that is already deployed in production,
   which makes it necessary to preserve the existing data.
   The desired system contains a number of changes and is meant to replace the existing system.

   Data migration for other purposes has been described in the literature.
   For instance, if a data migration is done for switching to another platform or to different technology,
   e.g.~\cite{Gholami2016,Bisbal1999},
   migration engineers may deliberately avoid schema differences and functionality changes to avoid introducing new errors in an otherwise error-prone migration process.
   For example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014} define an increment as a variation between two data structures.
   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
   Such data migrations do not allow for schema changes and do not fit our need for incremental software deployment.
   Examples like these have convinced the authors to define a new method for data migration that is specifically meant for systems in production that are
   deployed in increments.

   The contribution of this paper is to derive a migration schema from two artifacts: an existing system
   (which has its own schema and data set) and the specification of a desired system (which has a different schema and will ``import'' relevant parts of the existing data).   
   Our theory is based on the following assumptions and requirements:
\begin{itemize}
   \item {\em incremental deployment}\\The data migration is meant to deploy a software increment in production.
   \item {\em data pollution}\\The existing data set may be polluted, but it satisfies its schema.
   \item {\em human intervention}\\The data migration may require human interaction, which may take time.
   %\item {\em semantic continuity}\\the meaning of data must be preserved.
   \item {\em zero down-time}\\The business continues during the migration without interruption.
   \item {\em generative software development}\\A compiler exists to generate a system from a given schema.
\end{itemize}
   This paper focuses on the correctness of the migration.
   Efficiency of the migration is beyond the scope.

\section{Analysis}
\label{sct:Analysis}
\subsection{Information Systems}
   The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
   This section defines information systems, after which we define a migration procedure from one system to another.

   Every user has her own tasks and responsibilities and may work from different locations and on different moments.
   This collective use by multiple users serves a purpose which we will loosely call ``the business''.
   It is this business that depends on the semantics of the data to draw the right conclusions and carry out their tasks.
   %This paper uses semantic constraints to represent this meaning.

   Information systems are typically used by actors (both users and computers) who are distributed and work with data.
   As a consequence, the data in a system changes continually.
   The state of the system is represented by a data set, typically represented in a persistent store or database.
   Events that the system detects may cause the state to change.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.45]{figures/datamigration-Pre-migration.png}
   \end{center}
\caption{Anatomy of an information system}
\label{fig:pre-migration}
\end{figure}
   To keep our theory technology independent, data sets are assumed to contain triples.
   This makes our theory valid for any kind of database that triples can represent,
   including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
   An important aspect of the system's semantics are the constraints, sometimes known as invariants or integrity rules.
   %The purpose of these rules is to keep all constraints satisfied.

>>>>

\usepackage{hyperref}
%\usepackage{multicol}
%\usepackage{footmisc}
%\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amssymb,stmaryrd}
\usepackage{mathrsfs}
\usepackage{mathtools}
%\usepackage{amsthm}
\usepackage[english]{babel}
%\usepackage[official,right]{eurosym}
\selectlanguage{english}
\hyphenation{ExecEngine DevOps}
%\newtheorem{lemma}{Lemma}

@@

\title{Data Migration under a Changing Schema in Ampersand}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Sebastiaan Joosten\inst{1}\orcidID{0000-0002-6590-6220}\\ \and
Stef Joosten\inst{2,3}\orcidID{0000-0001-8308-0189}}
%
\authorrunning{S. Joosten}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{University of Minnesota, Minneapolis, USA \and Open Universiteit, Heerlen, the Netherlands\\ 
\email{stef.joosten@ou.nl} \and
Ordina NV, Nieuwegein, the Netherlands}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Software generators can help increase the frequency of releases and their reliability.
They save on time spent on development and fixing human-induced mistakes by compiling a specification into a working information system.
However, many generators do not support data migrations.
Data migration is necessary when an incremental deployment changes the system's schema.
Consequently, developers tend to avoid migrations or migrate data ``by hand''.

@@

\keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change.}
\end{abstract}
%
%
%
\section{Introduction}
\label{sct:Introduction}
   In practice, information systems%
   \footnote{In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.}
   may live for many years.
   After they are built, they need to be upgraded at an ever-increasing pace to keep up with changing requirements in a dynamically evolving environment.
   Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
   Obviously, these deployments are mostly upgrades of existing systems.

   A problem occurs when an upgrade involves a change in the schema of the deployed system:
   The production data of the existing system must be migrated to fit the schema of the migrated system.
   Let us call that a "schema changing data migration" (SCDM).
   For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
   Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
   For good reasons, most development teams will therefore try to avoid schema changes at all costs, fearing the risk and effort of the data migration.

   During the lifetime of a system, schema changing data migrations cannot always be avoided, however.
   It is this problem that we address in this paper.
   Data migration for other purposes has been described in the literature.
   For instance, if a data migration is done for switching to another platform or to different technology,
   e.g.~\cite{Gholami2016,Bisbal1999},
   migration engineers can and will avoid schema changes and functionality changes to avoid introducing new errors in an otherwise error-prone migration process.
   Another example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014}, defines a migration as a variation between two data structures.
   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
   This does not cater for schema changes, however.
   Then there are SCDMs in situations without a schema or an implicit schema, e.g.~\cite{Hillenbrand2022}.
   If the schema is not available at compile time, the work will have to be done at runtime.
   This leads to versioned storage and an overhead in performance.
   This paper focuses on SCDMs in situations with an explicit schema that changes with the data migration.
   The prototypical use case for that is to frequently release upgrades of information systems in production,
   where the semantic integrity of data must be preserved across schema changes.
   Another use case is application integration for multiple dispersed data sources with explicit schemas.

   To analyze this situation, we need to define the notion of schema and how it constrains the data.
   To facilitate a formal analysis of the situation,
   we define an information system (section~\ref{sct:Information Systems}) as a pair of a schema and a dataset,
   so that we have the schema explicitly at our disposal.
   The dataset is separate, so we may consider an analysis of the schema only as a compile-time activity.
   Section~\ref{sct:Schemas} formalizes schemas as a combination of concepts, relations and rules.
   Relations are carriers of the data and concepts are needed to define them.
   Rules serve as invariants on the dataset; they constrain the data to impose semantics on the data.
   For experimental validation we use the language Ampersand~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018}
   because its syntax and semantics correspond directly to these definitions.
   It has a notion of rule, which implements the invariants on the dataset.
   In this way, prototypes built in Ampersand have helped to refine the theory.

   In practice, data migrations typically follow Extract-Transform-Load (ETL) patterns~\cite{Theodorou2017},
   for which many tools are available.
   However, invariants that change yield extra work for software engineers to write the transform code,
   for which ETL tools typically provide little support.
   If automation can improve that, the frequency of SCDM's might well increase,
   and downtime for the sake of migration will become less acceptable.
   So, our research aims for zero-downtime as perceived by end-users.
   Another practical problem is that of data quality.
   Migrations typically suffer from a backlog of deteriorated data, incurring work to clean up data.
   Some of that must be done before the migration; some can wait till after the migration.
   We can define part of the data quality as satisfying semantic constraints.
   E.g. the constraint that the combination of street name, house number, postal code, and city occurs in
   a registration of valid addresses can be checked automatically.
   We can actually use invariants to help find and correct data pollution.
   We cannot use always use invariants, however.
   An example is when a person has deliberately specified a false name, without violating any of the existing invariants.

   To make the case for zero-downtime, we must distinguish between three types of constraint:
\begin{enumerate}
\item Transactional invariant\\
   A \define{transactional invariant} is a constraint is always true in a system.
   This is the classical database transaction; it may be violated temporarily, but the outside world will never notice.
   Thus, the zero-downtime requirement holds in the eyes of an end user.
\item Business invariant\\
   A \define{business invariant} is a constraint that that can be violated temporarily until the computer restores it.
   Example: All purchase orders are reflected on an order sheet, but different representations of the order sheet may be out of sync.
   On my smartphone, I see 7 purchase orders, but in my ERP system I see 6.
   The inconsistency arises because the back end application is still working on the consistency of the two.
   As a user, I must trust that the two will eventually be consistent.
   In this paper, we assume that this delay can be kept short enough to be negligible for users,
   so that the zero-downtime requirement is satisfied in their perception.
\item Business constraint\\
   A \define{business constraint} is a constraint that can be violated temporarily until a user restores it.
   Example: ``An authorized manager has to sign every purchase order.''
   We consider business constraints not to be invariants because users can violate them temporarily.
   This delay is definitely noticable for users.
   To satisfy the zero-downtime requirement, the system must be up and running to let users restore business constraints.
\end{enumerate}

Summarizing, the following requirements apply to SCDM'S
\begin{enumerate}
\item users must experience zero-downtime.
   Delays caused by transactional invariants or business invariants are not considered as downtime.
\item users must be able to fix business constraints, so they can be used
\item 
\end{enumerate}

   We can also use invariants to signal data pollution and make the migration system fix that.
   For that purpose we need to compute violations of invariants, so that we can restore them.
   The most important use of invariants, however, is to ensure that the migration system preserves the semantics of the data as much as possible.
   That prevents operational disruptions in the business caused by mistakes during the migration.

------------
oud

   This paper proposes a theory for data migrations that is meant to be implemented in a software generator called Ampersand.
   It aims at automating data migrations, so incremental changes can be done more reliably and faster.
   This contributes to a more agile software development process.
   We believe that users will experience a more organic evolution of the system because increments are smaller, occur more frequently and more reliably.

   We have not found any literature on schema changing data migrations.
   Examples like these have convinced the authors to define a new method for data migration that is specifically meant for systems in production that are
   deployed in increments.

   The contribution of this paper is to derive a migration schema from two artifacts: an existing system
   (which has its own schema and data set) and the specification of a desired system (which has a different schema and will ``import'' relevant parts of the existing data).   
   Our theory is based on the following assumptions and requirements:
\begin{itemize}
   \item {\em incremental deployment}\\The data migration is meant to deploy a software increment in production.
   \item {\em data pollution}\\The existing data set may be polluted, but it satisfies its schema.
   \item {\em human intervention}\\The data migration may require human interaction, which may take time.
   %\item {\em semantic continuity}\\the meaning of data must be preserved.
   \item {\em zero-downtime}\\The business continues during the migration without interruption.
   \item {\em generative software development}\\A compiler exists to generate a system from a given schema.
\end{itemize}
   This paper focuses on the correctness of the migration.
   Efficiency of the migration is beyond the scope.

\section{Analysis}
\label{sct:Analysis}
\subsection{Information Systems}
   This section defines information systems, so we can formally define the migration procedure from one system to another.

   The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
   Multiple users, working from different locations and on different moments, constitute what we will loosely call ``the business''.
   The data in the system constitutes the collective memory of the business,
   which relies on the semantics of the data to draw the right conclusions and carry out their tasks.
   %This paper uses semantic constraints to represent this meaning.

   Actors (both users and computers) are changing the data in a system continually.
   The state of the system is represented by a data set, typically represented in a persistent store or database.
   Events that the system detects may cause the state to change.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.45]{figures/datamigration-Pre-migration.png}
   \end{center}
\caption{Anatomy of an information system}
\label{fig:pre-migration}
\end{figure}
   To keep our theory technology independent, data sets are assumed to contain triples.
   This makes our theory valid for any kind of database that triples can represent,
   including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
   An important aspect of the system's semantics are the constraints, sometimes known as invariants or integrity rules.
   %The purpose of these rules is to keep all constraints satisfied.
